A Type System for Structured Aliasing in Parallel Programs

### 2012-10-30 doc/types-pldi2013/example.tex
Lots of small changes

The entry function is {\tt simulate\_circuit} (line 15), which is
parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn}.
The wires and nodes are contained in regions {\tt rw} and {\tt rn},
respectively.  The {\tt rl} region holds lists of pointers to the nodes and wires.
When {\tt simulate\_circuit} is invoked it must have both read and
write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
(line 17); the type system enforces that
{\tt simulate\_circuit} only accesses these regions and their
subregions at compile-time.

@@

After creating the colorings, the application partitions the node and wire
regions into {\em subregions} according to the colorings (lines 21-25).  For
simplicity, our example only partitions the circuit into two pieces.
Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
Subregion {\tt rn0} will contain all locations with a color of 1 in {\tt pc.1},
while {\tt rn1} will contain all locations with a color of 2.  This
partitioning uses a coloring and results in {\em disjoint} subpartitions.
Constraints are introduced into the static environment describing both the
disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
and only the inclusion constraints are introduced into the static environment.
Although multicolorings are not part of the Core Legion described in 
Section~\ref{sec:legioncore}, they can be easily implemented using a separate 
coloring for each color in the multicoloring.

@@

After partitioning the circuit into pieces, the application creates
instances of {\tt CircuitPiece} (defined on lines 9-12) for each piece
(lines 28-33).  {\tt CircuitPiece} is a {\em region relationship}, a
bounded existential type.  Region relationships allow the programmer
to {\em pack} a group of regions together and remember properties
about them such as disjointness and subregion relationships (lines 29
and 33).  The type system verifies the properties hold statically when
packing; when region relationships are {\em unpacked} the same
properties are reintroduced into the static checking environment
(lines 40-41).  A key feature underlying the soundness of the Legion
type system is that privileges cannot be packed in a region
relationship---privileges belong to functions. When a function unpacks
a region relationship it must already hold privileges for the new
regions it finds in the packed value.  For example, on line 40
function {\tt execute\_time\_steps} has read/write privileges for
region {\tt rn0} because it has read/write privileges for {\tt rn} and
the {\tt CircuitPiece} region relationship constraints ensure $\tt rn0
\leq rn$ (line 12).

>>>>

The entry function is {\tt simulate\_circuit} (line 15), which is
parameterized on regions {\tt rl}, {\tt rw}, and {\tt rn}.
The wires and nodes are contained in regions {\tt rw} and {\tt rn},
respectively.  The {\tt rl} region holds lists of pointers to the nodes and wires.
When {\tt simulate\_circuit} is invoked it must have both read and
write privileges for the regions {\tt rl}, {\tt rw}, and {\tt rn}
(line 17); the type system statically enforces the functional restriciton that
{\tt simulate\_circuit} only accesses these regions and their subregions.

@@

After creating the colorings, the application partitions the node and wire
regions into {\em subregions} according to the colorings (lines 21-25).  For
simplicity, our example only partitions the circuit into two pieces.
Line 21 uses coloring {\tt pc.1} to partition {\tt rn} into {\tt rn0}, and {\tt rn1}.
Subregion {\tt rn0} will contain all locations with a color of 0 in {\tt pc.1},
while {\tt rn1} will contain all locations with a color of 1.  This
partitioning uses a coloring and results in {\em disjoint} subpartitions.
Constraints are introduced into the static environment describing both the
disjointness of the subregions (i.e. {\tt rn0 $*$ rn1}) and the inclusion of
each subregion in the original region (e.g. {\tt rn0 $\leq$ rn}).  Line
23 is similar, but uses a multicoloring, so the subregions may be {\em aliased},
and only the inclusion constraints are introduced into the static environment.
Although multicolorings are not part of the Core Legion described in 
Section~\ref{sec:legioncore}, they can be easily implemented using a separate 
coloring for each color in the multicoloring.

@@

After partitioning the circuit into pieces, the application creates
instances of {\tt CircuitPiece} (defined on lines 9-12) for each piece
(lines 28-33).  {\tt CircuitPiece} is a {\em region relationship}, a
bounded existential type.  Region relationships allow the programmer
to {\em pack} a group of regions together and remember properties
about them such as disjointness and subregion relationships (lines 29
and 33).  The type system verifies the properties hold statically when
packing; when region relationships are {\em unpacked} the same
properties are reintroduced into the static checking environment
(lines 40-41).  A key feature underlying the soundness of the Legion
type system is that privileges cannot be packed in a region
relationship---privileges belong to functions. When a function unpacks
a region relationship it must already hold privileges for the new
regions it finds in the packed value.  For example, on line 40 the
function {\tt execute\_time\_steps} has read/write privileges for
region {\tt rn0} because it has read/write privileges for {\tt rn} and
the {\tt CircuitPiece} region relationship constraints ensure $\tt rn0
\leq rn$ (line 12).

### 2012-10-30 doc/types-pldi2013/intro.tex
Lots of small changes

These two approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single static partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach permits arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

In this work, we present a type system and operational semantics for
Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
dynamic partitions of program data which can result in aliasing.  We prove
the soundness of our type system and show that through a combination of
static and dynamic analysis we can implement it in a way that minimizes dynamic
runtime checks and results in low actual runtime overhead.  We then show 
that the soundness of our type system permits a scalable, hierarchical 
distributed scheduling algorithm that enables Legion programs to achieve
high performance on current supercomputers.

Legion presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that allows programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

To enforce this restriction, our type system relies on a combination of static 
and dynamic analyses.  The key observation is that dynamic
analysis is very easy and relatively cheap when done at the granularity of
regions instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

>>>>

These two approaches illustrate different mechanisms for dealing with
the problem of data aliasing in parallel programs.  The fully static approach disallows aliasing
by requiring a single disjoint partitioning of data.  Our own experience with writing high-performance 
applications in both Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple, simultaneous partitions of
the same data, providing different views on to that data.  Alternatively, 
the fully dynamic approach permits arbitrary data aliasing, but requires 
centralized control logic to perform conflict detection between every pair 
of parallel computations. This results in runtime overhead and is 
not scalable on distributed-memory machines.

In this work, we present a type system and operational semantics for
Legion\cite{Legion12}.  Legion is a programming model that supports multiple,
dynamic partitions of program data which can result in aliasing.  We prove
the soundness of our type system and show that through a combination of
static and dynamic analysis we can implement it in a way that minimizes dynamic
runtime checks and results in low actual runtime overhead.  We then show 
that the soundness of our type system permits a hierarchical,
distributed scheduling algorithm that enables Legion programs to scale to 
high performance on current supercomputers.

Legion presents {\em logical regions} as an
abstraction for capturing {\em locality} (data that will be used together,
and therefore should be colocated) and {\em independence} (disjoint data
that may be operated on in parallel).  Logical regions have the following
properties that allow programmers to dynamically communicate information
about the structure of program data to Legion:
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

To enforce this restriction, our type system relies on a combination of static 
and dynamic analyses.  The key observation is that dynamic
analysis is very easy and relatively cheap when done at the granularity of
regions instead of individual heap locations.  To that end, our type system supports
static verification of all pointer dereferences and passing of region privileges at 
task calls boundaries.  The Legion runtime then leverages its knowledge of logical
region aliasing to perform dynamic safety checks for the unpacking
and packing of logical regions to/from the heap with regards to task privileges.  
These checks are then amortized over all the uses of the logical region 
which significantly reduces their cost.

### 2012-10-30 doc/types-pldi2013/langdef.tex
Lots of small changes

Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
information about the overall environment.  The information consists of the initial state of the heap $S$, and a
set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
and is used to relax the constraints the evaluation of an expression.  Whereas the requirement for locations not
in the clobber set is the usual one of {\em consistency} (i.e. a read from a location observes the effect of all 
previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.

@@

Although determining that $e_1$
and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
non-interfering, the converse is not true.  It is possible for $e_1$
and $e_2$ to be judged to be potentially interfering and be serialized
even when the subexpressions are non-interfering.  This loss of
precision can impact performance, and did in a few instances in our
experiments.  
%When the performance impact is significant, two options
%exist to improve matters.  
In cases where $e_1$ and $e_2$ are in fact
non-interfering, region partitions can be further refined to make the
disjointness visible in the privileges.  Alternately, the code can be
refactored to lift the subexpressions up to the same level, allowing
the dynamic test based on the region mapping to infer non-interference
in more cases.  
%Although this can increase the overhead of scheduling
%decisions, judicious use of this technique in practice has not shown
%any measurable performance degradation.

>>>>

Instead we use a bottom-up approach with large-step semantics in which each expression is evaluated with limited 
information about the overall environment.  The information consists of the initial state of the heap $S$, and a
set of memory locations that might be modified by other concurrent evaluations.  This is the clobber set $C$,
and is used to relax the constraints on the evaluation of an expression.  Whereas the requirement for locations not
in the clobber set is the usual one of {\em coherence} (i.e. a read from a location observes the effect of all 
previous modifications applied in order), the value read from a location in the clobber set may be hypothesized
to be any value at all.  The validation of these hypotheses is deferred until all potentially-concurrent
modifications to a location can be observed (i.e. the evaluation of a parent expression whose clobber set does
not include the location) and a consistent interleaving of the subexpressions' memory traces can be determined.
With this formulation, the criteria for when a hierarchical scheduler may execute subexpressions in parallel can be
made clear, but they are no longer constructive.  Instead, we will have to show that the scheduling algorithm
used by the Legion runtime satisfies these constraints, guaranteeing the soundness of the Legion programming model.

@@

Although determining that $e_1$
and $e_2$ are non-interfering guarantees $e_3$ and $e_4$ are
non-interfering, the converse is not true.  It is possible for $e_1$
and $e_2$ to be judged to be interfering and be executed sequentially 
even when the subexpressions are non-interfering.  This loss of
precision can impact performance, and did in a few instances in our
experiments.  
%When the performance impact is significant, two options
%exist to improve matters.  
In cases where $e_1$ and $e_2$ are in fact
non-interfering, region partitions can be further refined to make the
disjointness visible in the privileges.  Alternately, the code can be
refactored to lift the subexpressions up to the same level, allowing
the dynamic test based on the region mapping to infer non-interference
in more cases.  
%Although this can increase the overhead of scheduling
%decisions, judicious use of this technique in practice has not shown
%any measurable performance degradation.
